---
title: "metadata_preparation"
author: "Leire A. Murua"
date: "2025-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

library(tidyverse); packageVersion("tidyverse")
library(dplyr); packageVersion("dplyr")

library(beepr)

```

## Load metadata files and combine them, modify for the data needed
## Add factorized column of disease severity according to calprotectin values

```{r}
metadata_737472 <- read.csv("../Data/Metadata/737472_SraRunTable.csv", row.names = 1)
metadata_951422 <- read.csv("../Data/Metadata/951422_SraRunTable.csv", row.names = 1)

# View(metadata_737472)
# names(metadata_737472)

metadata_737472_1 <- metadata_737472 %>% 
  select(BioProject, Sample.Name, host_is_fmt_recipient, 
         host_subject_id, collection_days_post_initial_fmt, Host_age,
         host_assigned_donor_subject_id, host_sex, host_subject_assigned_to_antibiotics_arm,
         host_subject_assigned_to_maintenance_arm, rrs_library_id..run., AGE, 
         Library_ID, Patient_ID, sex, Donor) %>% 
  rename_all(tolower) %>% 
  mutate(host_subject_id = case_when(
    patient_id == "DFS.007" ~ "S0007",
    TRUE ~ host_subject_id
  )) %>% 
  unite("ttm_arm", host_subject_assigned_to_antibiotics_arm, 
        host_subject_assigned_to_maintenance_arm, sep = "/") %>% 
  mutate(host_assigned_donor_subject_id = case_when(
    grepl("^D", host_subject_id) ~ "is donor",
    TRUE ~ host_assigned_donor_subject_id 
  )) %>% 
  mutate(ttm_arm = case_when(
    host_assigned_donor_subject_id == "is donor" ~ "healthy",
    host_subject_id == "S0007" ~ "ABX-/CAPS",
    TRUE ~ ttm_arm
  )) %>%
  mutate(coll_time = case_when(
    grepl("^D", host_subject_id) ~ "",
    grepl("^S", host_subject_id) ~ substr(rrs_library_id..run., start = 9, stop = nchar(rrs_library_id..run.)),)) %>% 
  select(-rrs_library_id..run., patient_id) %>% 
  mutate(subject_type = case_when(
    donor == "yes" ~ "Healthy", 
    host_is_fmt_recipient == "True" ~ "UC",
    grepl("^S", host_subject_id) ~ "UC",        # Assign "UC" to samples with host_subject_id starting with "S"
    grepl("^D", host_subject_id) ~ "Healthy"    # Assign "Healthy" to samples with host_subject_id starting with "D"
  )) %>% 
  mutate(disease_status_end = case_when(
    subject_type == "Healthy" ~ "healthy",
    host_subject_id == "S0001" ~ "active", 
    host_subject_id == "S0004" ~ "remission",
    host_subject_id == "S0007" ~ "active",
    host_subject_id == "S0008" ~ "remission",
    host_subject_id == "S0013" ~ "active",
    host_subject_id == "S0017" ~ "active",
    host_subject_id == "S0021" ~ "active",
    host_subject_id == "S0024" ~ "active",
    host_subject_id == "S0027" ~ "remission",
    host_subject_id == "S0037" ~ "active",
    host_subject_id == "S0041" ~ "active",
    host_subject_id == "S0043" ~ "active",
    host_subject_id == "S0047" ~ "remission",
    host_subject_id == "S0053" ~ "remission",
    host_subject_id == "S0055" ~ "active",
    host_subject_id == "S0056" ~ "active",
    host_subject_id == "S0059" ~ "active",
  )) %>% 
  mutate(donor = case_when(
    host_assigned_donor_subject_id == "is donor" ~ "yes",
    host_is_fmt_recipient == "True" ~ "no",
    TRUE ~ donor
  )) %>% 
  filter(!host_subject_id %in% c("S0060", "S0061", "S0062", "S0063", "S0064")) %>% 
  select(bioproject, sample.name, host_subject_id, 
         host_is_fmt_recipient, host_assigned_donor_subject_id, 
         collection_days_post_initial_fmt, coll_time, ttm_arm, host_age, host_sex,
         age, sex, library_id, donor, subject_type, disease_status_end) %>% 
  mutate(collection_days_post_initial_fmt = case_when(
    host_subject_id %in% c("D0044", "D0097", "D0485") ~ NA_real_,
    TRUE ~ collection_days_post_initial_fmt
  )) %>% 
  mutate(coll_time = case_when(
    host_subject_id %in% c("D0044", "D0097", "D0485") ~ "not app",
    TRUE ~ coll_time
  )) %>% 
  mutate(host_sex = case_when(
    host_subject_id %in% c("D0044", "D0097", "D0485") ~ "not app",
    host_subject_id == "S0007" ~ "male", 
    TRUE ~ host_sex
  )) %>% 
  mutate(host_age = case_when(
    host_subject_id %in% c("D0044", "D0097", "D0485") ~ NA_real_,
    host_subject_id == "S0007" ~ 34, 
    TRUE ~ host_age
  )) %>% 
  mutate(host_is_fmt_recipient = case_when(
    host_subject_id == "S0007" ~ "True", 
    TRUE ~ host_is_fmt_recipient
  )) %>% 
  mutate(host_assigned_donor_subject_id = case_when(
    host_subject_id == "S0007" ~ "D0485", 
    TRUE ~ host_assigned_donor_subject_id
  )) %>% 
  select(-c(age, sex, library_id)) %>% 
  filter(host_subject_id != "") %>% 
  filter(!host_subject_id %in% c("S0017", "S0043", "S0037", "S0031", "S0029")) %>% 
  rownames_to_column(var="samples") %>% 
  filter(grepl("^SRR15", samples)) %>% 
  column_to_rownames(var="samples")

beep(sound=2)

``` 



```{r}

metadata_737472_1 <- metadata_737472_1 %>% 
  mutate(ttm_arm = case_when(
    ttm_arm == "/" ~ "no data",
    TRUE ~ ttm_arm 
  )) %>% 
  mutate(timepoint = collection_days_post_initial_fmt)


saveRDS(metadata_737472_1, "metadata_737472_1.rds")
names(metadata_737472_1)

beep(sound=2)
```

## Only keep "disease_status_end" value for last collection of each subject.

```{r}

rownames_df_1 <- rownames(metadata_737472_1)

metadata_737472_1 <- metadata_737472_1 %>%
  group_by(host_subject_id) %>%
  mutate(
    max_collection_day = max(collection_days_post_initial_fmt, na.rm = TRUE),
    disease_status_end = ifelse(
      disease_status_end != "healthy" & collection_days_post_initial_fmt != max_collection_day, 
      NA, 
      disease_status_end
    )
  ) %>%
  ungroup() %>% 
  select(-max_collection_day) %>% 
  rownames_to_column(var="samples") %>% 
  mutate(timepoint = as.numeric(timepoint)) %>%
  group_by(host_subject_id) %>%
  mutate(
    baseline_status = case_when(
      timepoint == min(timepoint, na.rm = TRUE) ~ "active", # Assign "active" to the row with the lowest timepoint
      grepl("^D", host_subject_id) ~ "healthy", 
      TRUE ~ NA_character_  # Assign NA to other rows 
    )
  ) %>% 
  group_by(host_subject_id) %>%
  mutate(timepoint = timepoint - min(timepoint)) %>%
  ungroup()

metadata_737472_1 <- as.data.frame(metadata_737472_1)

rownames(metadata_737472_1) <- rownames_df_1 

beep(sound=2)
```



```{r}
# View(metadata_951422)
names(metadata_951422)

metadata_951422_1 <- metadata_951422 %>%
  select(BioProject, Sample.Name, host_subject_id, Host_age, 
         host_body_mass_index, host_sex, Sample.Name, timepoint, 
         fecal_calprotectin) %>% 
  rename_all(tolower) %>% 
  mutate(timepoint = tolower(timepoint)) %>% 
  mutate(disease_status = case_when(
    timepoint == "flare" ~ "active",
    timepoint == "subjective_flare_but_did_not_meet_criteria_for_clinical_flare" ~ "active",
    TRUE ~ "remission"
  )) 

saveRDS(metadata_951422_1, "metadata_951422_1.rds")
# View(metadata_951422_1)

beep(sound=2)
```

```{r}

# metadata_737472_1 <- readRDS("metadata_737472_1.rds")
# metadata_951422_1 <- readRDS("metadata_951422_1.rds")

dim(metadata_737472_1)
dim(metadata_951422_1)

names(metadata_737472_1)
names(metadata_951422_1)

met_737472_comb <- cbind(metadata_737472_1, host_body_mass_index=NA, 
                         fecal_calprotectin=NA, disease_status=NA)

met_951422_comb <- cbind(metadata_951422_1, host_is_fmt_recipient="False", 
                         coll_time=NA, 
                         host_assigned_donor_subject_id="not app", ttm_arm="none", 
                         donor="no", subject_type="UC",
                         disease_status_end=NA) 

met_737472_comb <- met_737472_comb %>% 
  select(bioproject, sample.name, host_subject_id, 
         host_is_fmt_recipient, host_assigned_donor_subject_id, coll_time, 
         timepoint, ttm_arm, host_age, host_sex, host_body_mass_index, donor, 
         subject_type, fecal_calprotectin, 
         baseline_status, disease_status, disease_status_end)

met_951422_comb <- met_951422_comb %>% 
  select(bioproject, sample.name, host_subject_id, 
         host_is_fmt_recipient, host_assigned_donor_subject_id, coll_time,
         timepoint, ttm_arm, host_age, host_sex, host_body_mass_index, donor, 
         subject_type, fecal_calprotectin, disease_status, 
         disease_status_end) %>% 
   mutate(timepoint = case_when(
    grepl(" mo$", timepoint) ~ as.character(as.numeric(gsub(" mo", "", timepoint)) * 30),
    TRUE ~ timepoint
  ))

## Fix 737472 metadata so only the "end" results show on the "disease_status_end" column, and the rest are turned to NA

rownames_df_2 <- rownames(met_737472_comb)


met_737472_comb <- as.data.frame(met_737472_comb)

rownames(met_737472_comb) <- rownames_df_2


dim(met_737472_comb)
dim(met_951422_comb)

names(met_737472_comb)
names(met_951422_comb)

beep(sound=2)
```

## Find disease status for end timepoint for each subject and add it to disease_status_end column

```{r}
rownames_df_2 <- rownames(met_951422_comb)

valid_timepoints <- c("90", "180", "270", "360", "450", "540")

met_951422_comb <- met_951422_comb %>% 
  select(host_subject_id, timepoint, disease_status, disease_status_end, everything()) %>% 
  rownames_to_column(var="samples") %>% 
  filter(!grepl("_cross-sectional$", sample.name)) %>% 
  group_by(host_subject_id) %>% 
  mutate(timepoint = case_when(
    timepoint == "baseline" ~ "0",
    timepoint %in% c("flare", "subjective_flare_but_did_not_meet_criteria_for_clinical_flare") ~ {
      numeric_timepoints <- as.numeric(timepoint[!timepoint %in% c("flare", "subjective_flare_but_did_not_meet_criteria_for_clinical_flare", "baseline")])
      if(length(numeric_timepoints) > 0) {
        as.character(max(numeric_timepoints, na.rm = TRUE) + 90)
      } else {
        "90"  
      }
    },
    TRUE ~ as.character(timepoint))) %>% 
  mutate(timepoint = as.numeric(timepoint)) %>% 
  mutate(
    baseline_status = case_when(
      timepoint == min(timepoint, na.rm = TRUE) ~ disease_status, # Assign disease_status to the row with the lowest timepoint
      TRUE ~ NA_character_  # Assign NA to other rows 
    ),
    disease_status_end = case_when(
      timepoint == max(timepoint, na.rm = TRUE) ~ disease_status, 
      TRUE ~ NA_character_   
    )
  ) %>%
  ungroup() %>% 
  column_to_rownames(var="samples") %>% 
  select(host_subject_id, timepoint, disease_status, baseline_status, disease_status_end, everything()) 


beep(sound=2)
```

```{r}
# Process data while keeping row names intact
# met_951422_comb <- met_951422_comb %>%
#   rownames_to_column(var="samples") %>% 
#   group_by(host_subject_id) %>%
#   mutate(
#     has_flare = any(timepoint == c("flare", "possible flare")),
#     highest_existing_timepoint = max(as.numeric(timepoint[timepoint %in% valid_timepoints]), na.rm = TRUE),
#     timepoint_count = n()  # Count number of timepoints per host_subject_id
#   ) %>%
#     mutate(timepoint = case_when(
#     timepoint == "baseline" ~ "0",
#     timepoint %in% c("flare", "subjective_flare_but_did_not_meet_criteria_for_clinical_flare") ~ {
#       numeric_timepoints <- as.numeric(timepoint[!timepoint %in% c("flare", "subjective_flare_but_did_not_meet_criteria_for_clinical_flare", "baseline")])
#       if(length(numeric_timepoints) > 0) {
#         as.character(max(numeric_timepoints, na.rm = TRUE) + 90)
#       } else {
#         "90"  # Default to 90 if no numeric timepoints are found
#       }
#     },
#     TRUE ~ as.character(timepoint)
#   )) %>% 
#   ungroup() %>%
#   filter(!grepl("_cross-sectional$", sample.name)) %>% 
#   mutate(
#     disease_status_end = case_when(
#       has_flare & timepoint == "flare" ~ disease_status,  # If "Flare" exists, copy its disease_status
#       !has_flare & timepoint == "540" ~ disease_status,   # If no "Flare", copy disease_status from "540"
#       !has_flare & is.na(highest_existing_timepoint) ~ disease_status_end,  # If no valid timepoints, do nothing
#       !has_flare & as.numeric(timepoint) == highest_existing_timepoint ~ disease_status,  # Copy from highest available timepoint
#       timepoint_count == 1 ~ disease_status,  # If only one timepoint exists for host_subject_id, copy its disease_status
#       TRUE ~ disease_status_end  # Keep existing values
#     )
#   ) %>%
#   mutate(baseline_status = "remission") %>% 
#   select(-has_flare, -highest_existing_timepoint, -timepoint_count) %>%  # Remove helper columns
#   group_by(host_subject_id) %>% 
#   
#   ungroup() %>% 
#   column_to_rownames(var="samples")
  
met_951422_comb$timepoint <- as.numeric(met_951422_comb$timepoint)


# Convert to data frame
met_951422_comb <- as.data.frame(met_951422_comb)


```


## Check how many samples of the same timepoint are taken for each subject.
```{r}
# # Count occurrences of each (host_subject_id, timepoint) combination
# duplicates <- met_951422_comb %>%
#   count(host_subject_id, timepoint) %>%
#   filter(n > 1)  # Keep only duplicates (n > 1)
# 
# # Print the duplicate combinations
# print(duplicates)
# 
# # Count total number of duplicate cases
# total_duplicates <- sum(duplicates$n) - nrow(duplicates)  # Exclude one occurrence per duplicate group
# cat("Total duplicate cases:", total_duplicates, "\n")
# 
# beep(sound=2)
```



```{r}

dim(met_737472_comb)
dim(met_951422_comb)
colnames(met_737472_comb)
colnames(met_951422_comb)

saveRDS(met_737472_comb, "met_737472_comb.rds")
saveRDS(met_951422_comb, "met_951422_comb.rds")

metadata_merged <- rbind(met_737472_comb, met_951422_comb)


metadata_merged <- metadata_merged %>% 
  mutate(bioproject = case_when(
    bioproject == "PRJNA737472" ~ 1,
    bioproject == "PRJNA951422" ~ 2 
  )) %>% 
  mutate(disease_status = case_when(
    subject_type == "Healthy" ~ "healthy",
    TRUE ~ disease_status
  ))


saveRDS(metadata_merged, "metadata_merged.rds")
beep(sound=2)
```


```{r}
metadata_merged <- metadata_merged %>% 
  rownames_to_column(var="samples") %>% 
  group_by(host_subject_id) %>% 
  mutate(group = paste(na.omit(baseline_status), na.omit(disease_status_end), sep="-")) %>% 
  filter(n() > 1) %>% 
  ungroup() %>% 
  column_to_rownames(var="samples") 
  
meta_merged <- metadata_merged %>% 
  mutate(timepoint = as.numeric(timepoint)) %>% 
  mutate(week = floor(timepoint/7)) %>% 
  mutate(week = factor(week))


saveRDS(meta_merged, "meta_merged.rds")

```

